A molecular taxonomy of tumors independent of tissue-of-origin

### 2020-08-05 text/manuscript.Rmd

Pathway analysis provides additional information that genes and gene ontologies
alone lack: 1) it aggregates molecular events across multiple genes in the same
pathway and 2) the results, interpretable as genetic alterations, are related 
to molecular functions such as “intrinsic pathway for apoptosis” or “cellular 
response to hypoxia”. Using the Reactome Pathway Database, we selected the 
second sub-pathways for each pathway (e.g. ‘Beta-catenin independent WNT 
signaling’ in ‘Signal Transduction’) which we believed was a suitable 
representation of the basic cellular processes and biochemical pathways. 
We also excluded pathways categorized under cancer, disease, gene defect, 
disorder, infection, reproduction, and muscle contraction (e.g. ‘PI3K/AKT 
Signaling in Cancer’). While some of the excluded pathways have been shown
to play an important role in the cancer, they focus on a specific point 
mutations. In addition, for most of the excluded pathways, there is a global
pathway (e.g. ‘PIP3 activates Akt signaling’) that exist. Finally, we mapped
the 18,577 Ensembl IDs from the GDC dataset to the highly selected pathways.
This operation produced a lookup table that consisted of 377 pathways mapped 
to 9,910 genes.

>>>>

To understand the molecular mechanism of cancer at a pathway level, our pathway
analysis approach focused on Reactome (https://reactome.org/), a pathway knowledge-based 
database. The mapping files of ENSEMBL to pathways, pathway hierarchy relationship,
and complexes to top pathways were downloaded from the download page 
(https://reactome.org/download-data). Using the data, we imposed pathway criteria to
meet the basic cellular processes and biochemical pathways: (1) human-derived pathways
(_“HSA”_) (2) limited to grandchild node for each parent pathway (_e.g._ ‘Beta-catenin 
independent WNT signaling’ in ‘Signal Transduction’) (3) exclusion of pathways in the
parent pathway: Disease, Muscle contraction, and Reproduction or pathway names that 
include any of the following keywords: disease, defect, cancer, oncogenic, disorder, 
mutant, loss, infection, bacteria, or listeria. While some of the excluded pathways 
have been shown to play an important role in cancer, they are highly specialized 
(_e.g._ ‘PI3K/AKT Signaling in Cancer’). Additionally, for most of the excluded pathways,
a neutral version pathway of the pathway exist (_e.g._ ‘PIP3 activates Akt signaling’). 
Finally, we mapped the 18,577 Ensembl IDs from the TCGA dataset to the highly selected
Reactome pathways. This operation produced a lookup table that consisted of 377 pathways
mapped to 8,940 genes.
